---
title: "Text and Sentiment Analysis"
description: |
  In this project, I used text wrangling techniques and sentiment analysis in R to analyze East of Eden, by John Steinbeck.
output: distill::distill_article
---

East of Eden is one of my favorite books, so undertaking a short text analysis project on this book was only natural. I used the `stringr` and `tidytext` packages to wrangle the text from a PDF of the book to get information on word count by chapter. This project was a lot more difficult for me as learning to work with strings is a lot different than working with numerical data. After getting word counts, I went into sentiment analysis and used the [NRC lexicon](https://saifmohammad.com/WebPages/NRC-Emotion-Lexicon.htm#:~:text=The%20NRC%20Emotion%20Lexicon%20is,were%20manually%20done%20by%20crowdsourcing.) to analyze the emotions associated with each word in the novel to get an overall feel of the text.

**Citation:** 

Steinbeck, J. (1992). East of Eden. New York: Penguin Books.

```{r setup, include = TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(tidytext)
library(textdata)
library(pdftools)
library(ggwordcloud)
library(kableExtra)
library(plotly)
```

```{r}
eoe <- pdf_text(here::here('data', '20211110-a5.pdf'))
```

### Word Counts

```{r}
eoe_text <- data.frame(eoe) %>% ### converts text to data frame
  mutate(page = 1:n(), ### add page number
         full_text = str_split(eoe, pattern = '\\n')) %>% ### incorporate line breaks
  unnest(full_text) %>% ### make each line a row
  mutate(full_text = str_squish(full_text)) %>% ### remove extra white space
  select(!1) %>% ### remove nested original text
  filter(page >= 6 & page <= 638) ### only include information in chapters
```

```{r}
eoe_chapts <- eoe_text %>% 
  mutate(chapter = ifelse(str_detect(full_text, "Chapter"), full_text, NA)) %>% ### detects chapter
  fill(chapter, .direction = 'down') %>% ### adds the chapter to every row
  separate(col = chapter, into = c("ch", "no"), sep = " ") %>% ### splits chapter and number
  mutate(no = ifelse(no == 'I', 1, no), ### roman numeral to numerical
         no = as.numeric(no)) %>% ### converts number to numeric class
  drop_na()
```

```{r}
eoe_words <- eoe_chapts %>% 
  unnest_tokens(word, full_text) %>% ### each row is a word   
  mutate(word = str_remove_all(word, pattern = "[:digit:]")) ### removes numerical words

eoe_word_count <- eoe_words %>% 
  group_by(no) %>% 
  summarize(word_count = n()) ### word count by chapter

ggplotly(ggplot(data = eoe_word_count, aes(x = no, y = word_count)) +
  geom_col(stat = 'identity', fill = 'darkblue', aes(text = paste('Chapter', no, '\nWord Count:', word_count))) +
    labs(x = 'Chapter',
         y = 'Word Count') +
  theme_minimal(),
  tooltip = 'text')
```

**Figure 1:** Word Count by Chapter for East of Eden

```{r}
eoe_words_clean <- eoe_words %>% 
  anti_join(stop_words, by = 'word') ### removes stop words from the list   

eoe_count <- eoe_words_clean %>% 
  count(word) %>% ### overall word count 
  slice(1:200) ### first 200 words

### wordcloud 
ggplot(data = eoe_count, aes(label = word)) +
  geom_text_wordcloud(aes(color = n, size = n), shape = 'triangle-upright', area_corr_power = 1) +
  scale_size_area(max_size = 24) +
  scale_color_viridis_c() +
  theme_minimal()
```

**Figure 2:** Word Cloud for the Top 200 most common words in East of Eden

The most common words were found to be character names. Other common words were found to be words describing time and ambiance, such as afternoon and air. 

### Sentiment Analysis

```{r}
### join the NRC Lexicon
eoe_sentiment <- eoe_words_clean %>% 
  inner_join(get_sentiments(lexicon = 'nrc'), by = 'word')
```

```{r}
nrc_count <- eoe_sentiment %>% 
  group_by(sentiment) %>% 
  summarize(count = n()) ### get counts of sentiments


ggplotly(ggplot(nrc_count) +
  geom_col(aes(x = reorder(sentiment, -count), y = count,
               text = paste('sentiment:', sentiment, '\ncount:', count)), fill = 'darkred') +
  theme_minimal() +
  labs(x = 'sentiment',
       y = 'count'),
  tooltip = 'text')
```

**Figure 3:** Sentiment Analysis using the NRC lexicon for East of Eden.

Interestingly, the text is almost equally positive and negative. East of Eden has many positive aspects to it, especially when describing the beauty of the Salinas Valley, but also contains many negative sentiments dealing with relationships and human feelings. Averaging out the two, the text is relatively neutral.


